---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
---

Set global parameters

``` {r setup}
library(knitr)
opts_chunk$set(cache=FALSE, fig.align='center')
```
## Loading and preprocessing the data

Unzip the file 

```{r Loading the File}

filename <- c("activity.csv")

if (!file.exists(filename)) {
    unzip("activity.zip")    
}

```

Read the file into memory

``` {r Read File}

data <- read.csv(filename, na.strings = "NA")

```

Convert date column into the Date format

``` {r Convert date column into the Date format}

data$date <- as.Date(data$date)

```


## What is the mean total number of steps taken per day?

Calculate total number of steps taken per day

``` {r Calculating total steps taken per day}

totalStepsPerDay <- tapply(data$steps, data$date, sum, na.rm = TRUE)

```

Make histogram of total steps taken each day

``` {r Histogram of total steps taken each day}

hist(totalStepsPerDay, 
     main = "Histogram of Steps Taken Per Day", 
     xlab = "Steps Taken Per Day")

```

Mean number of steps per day

``` {r Mean number of steps per day}

mean(totalStepsPerDay)

```

Median number of steps per day

``` {r Median number of steps per day}

median(totalStepsPerDay)

```

## What is the average daily activity pattern?

Calculate average steps taken per interval

``` {r Average Steps Per Interval}

avgStepsPerInterval <- tapply(data$steps, data$interval, mean, na.rm = TRUE)

```

Reformat and sort data

``` {r Reprocessing}

# Convert to a data frame
df <- data.frame(avgStepsPerInterval)

# Make the row names their own column
df <- cbind(interval = rownames(df), df)

# Remove rownames
rownames(df) <- NULL

# Convert intervals to integers
df$interval <- as.integer(as.character(df$interval))

# Sort by increasing interval order
df <- df[order(df$interval), ]

```

Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

``` {r Time Series Plot}

title <- c("Average Number of Steps Taken Per Interval")
ylabel <- c("Average Number of Steps Taken")
xlabel <- c("Interval (5 minutes)")

plot(x = df$interval, y = df$avgStepsPerInterval, type = "l", main = title, ylab = ylabel, xlab = xlabel)

```

```{r Max Steps Taken in Interval}

# Find interval with maximum number of steps taken
sortedDescendingStepsPerInterval <- df[order(df$avgStepsPerInterval, decreasing = TRUE), ]
sortedDescendingStepsPerInterval[1,1]

```

## Imputing missing values

Total number of missing values

``` {r Calculate number of missing values}

# Count total number of missing values
missing <- which(is.na(data$steps))
length(missing)

```

Imputing Missing Values

```{r Imputing Missing Values}

# Create new dataset to include imputed missing values
dataWithImputedValues <- data

# Function to fill in missing values
fillin <- function() {
  
    # for each row of missing data, get and insert the average for the interval
    for (rowIndex in missing) {
  
        # get interval
        interval <- dataWithImputedValues[rowIndex, "interval"]
      
        # get interval average
        intervalAverage <- df[df$interval == interval, "avgStepsPerInterval"]
      
        # replace missing value with interval average
        dataWithImputedValues[rowIndex, "steps"] <<- intervalAverage
    }
}

# Actually fill in missing values
fillin()

```

Check again for missing values - should be 0!

``` {r Check again for missing values}

nrow(dataWithImputedValues[is.na(dataWithImputedValues$steps),])

```

Recalculate total number of steps per day. This time, it includes imputed values

``` {r Recalculate total number of of steps per day}

# Calculate total number of steps taken per day
totalStepsPerDay <- tapply(dataWithImputedValues$steps, 
                           dataWithImputedValues$date, 
                           sum, 
                           na.rm = TRUE)

```

Another histogram of total steps taken each day. This time, it includes imputed values

``` {r Histogram with imputed values}

hist(totalStepsPerDay, 
     main = "Histogram of Steps Taken Per Day", 
     xlab = "Steps Taken Per Day")

```

Mean number of steps per day (with imputed values)

``` {r Mean with imputed values}

mean(totalStepsPerDay)

```

Median number of steps per day (with imputed values)

``` {r Median with imputed values}

median(totalStepsPerDay)

```

#### What is the impact on the mean and median with imputed values?

The mean and median of the steps taken per day are greater when we impute missing values than when we excluded them in prior calculations. Imputing missing values increases the estimate of the total monthly number of steps.

## Are there differences in activity patterns between weekdays and weekends?

Load the data.table library and convert data frame to a data table

``` {r load data.table}
library(data.table)

dt <- data.table(dataWithImputedValues)
```

Create new column for day of the week

``` {r Create new column for day of the week}

dt[,dayofweek:=weekdays(date)]

```

Create column for weekday status and converge into a 2-level factor

``` {r Create column for weekday status}

dt[,weekdayStatus:=ifelse(dayofweek %in% c("Saturday", "Sunday"), c("weekend"), c("weekday"))]

# Converge to a 2-level factor variable
dt$weekdayStatus <- factor(dt$weekdayStatus)

```

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

Panel plot setup

``` {r Panel Plot Setup}

# For panel plots, let's use the lattice plotting system
library(lattice)

# Set up plot parameters
title <- c("Average Number of Steps Taken Per Interval")
ylabel <- c("Average Number of Steps Taken")
xlabel <- c("Interval (5 minutes)")

``` 

Format data for plotting

```{r Format Data For Plotting}

plotData <- dt[,list(avgSteps = mean(steps)), by = c("interval", "weekdayStatus")]

```

Make panel plot

``` {r Make Panel Plot}
xyplot(avgSteps ~ interval | weekdayStatus, 
       data = plotData, 
       layout = c(1,2), 
       type = "l", 
       xlab = xlabel,
       ylab = ylabel,
       main = paste(title, c("By Weekday Status"), sep = " "))

```
